---
title: Using cbsodataR to plot total deaths per week in the Netherlands from 1995
  to 2020 (OC from reddit)
summary: "Replicating an original content (OC) from reddit with the R packages cbsodataR and ggplot2."
author: admin
date: '2020-05-17'
slug: cbsodatar-reddit
categories: []
featured: true
tags:
  - R
  - cbsodataR
  - ggplot2
image:
  caption: "test"
---

# Background
The other day I was browsing through reddit when a post in **r/dataisbeautiful**
caught my attention. The title stated
[[OC] Total deaths per week in the Netherlands from 1995 to 2020](https://www.reddit.com/r/dataisbeautiful/comments/gcqqfn/oc_total_deaths_per_week_in_the_netherlands_from/) and showed a graph with several lines. User *theorange1990*
plotted the weekly total deaths in
the Netherlands between 1995 and 2020, together with an "Upper Bound"
line. In addition, three more labels were included in the figure, pointing to
"2018 - Flu epidemic in the Netherlands", "2020 - so far", and '"Upper Bound" of
a normal year curve'.

What stood out was one yellow curve, peaking on the y-axis. This one was
labelled with "2020 - so far". It didn't look long for the community to point to
some methodological flaws of plotting the data "as is", but I was interested in
where the data were actually coming from. In the authors citation, *theorange1990*
mentioned the source to be Statistics Netherlands (CBS), the national statistical
office of the Netherlands and just below the sentence: "Data organized and 
graphed using Excel".

While Excel is of course a valid tool for such simple plotting exercises, I
wanted to see if and how I can replicate this graph in R.

The replication took about 10-15 minutes with downloading the most
recent data from the CBS servers directly into R. This approach has several
advantages:

- R is getting newest data directly from the CBS server (no separate downloading,
saving, opening, transformation from .csv to .xlsx by hand necessary)
- The code can easily be shared (what I do here) and everyone running the script
will received the same result
- With new data points becoming available, the plot is "updating itself"

In the following, I want share my approach.

# Needed packages

- **tidyverse** for data cleaning, transformation and visualisation
- **cbsodataR** open data API client for CBS

# CBS data

## Download with CBS API: cbsodataR
To download the same data as the reddit author, we look at his/her link
provided in the post: https://opendata.cbs.nl/#/CBS/en/dataset/70895ENG/table

Clicking the link brings us to the online table but we just need the "table
identifier" which is a unique table ID for each CBS tables. I found these difficult
to find on the table website but if you have the link, the identifier is part of
its URL. In this case this is `70895ENG`.

Let's start writing the R script.

```{r echo=FALSE, message=FALSE}
# Load needed packages
library(tidyverse)
library(cbsodataR)

# Import data from CBS using the table identifier
db_cbs <- cbsodataR::cbs_get_data("70895ENG") %>%
        cbs_add_label_columns() # Without this line we only see codes for each observation which are difficult to interpret

```

Now that we have downloaded the table, let's have a look at it.

```{r}
head(db_cbs)
```

## Tidy data
The formatting looks different from the format on the table's website, but this
doesn't matter for this plotting exercise.

One problem we have with the table though is that year and weeks are "smashed
together" into one variable: `Periods` or `Periods_label`.

We however want to plot deaths by week on the x-axis and each year should become
one line. So we need to separate the years from the weeks.

Since `Periods` always has eight characters, it is easiest use this column and
separate the two information on time with `substr()`.

Just as on the OC we are only interested in pooled data for:

- both sexes (see columns `Sex` or `Sex_label`),
- all age groups (see columns `Age31Decembet` or `Age31December_label`), and
- weeks 1 to 52 (weeks 0 and 53 were excluded since they do not appear in each year).

```{r}
db_plot <- db_cbs %>%
  mutate(year = factor(substr(Periods, start = 1, stop = 4)), # extract year
         wk = factor(as.numeric(substr(Periods, start = 7, stop = 8)))) %>% #extract weeks
        filter(Sex == "T001038", # only keep data for both sexes
               Age31December == "10000",
               wk %in% 1:52) %>% # only keep weeks 1 through 52
        select(Sex_label, Deaths_1, year, wk) %>% # keep necessary data only
  mutate(Sex_label = factor(Sex_label))
```

Now that we have the data in a tidy way and kept only necessary columns, let's
have a look at it (first six rows should be enough).

```{r}
head(db_plot)
```

## Plotting data

In this exercise we will use a simple line chart with **ggplot2**. Since we have
tidy data already we just needs to specify a few things.

```{r plotting}
ggplot(data = db_plot) +
  # plot lines
  geom_line(aes(x = factor(wk), y = Deaths_1, col = year, group = year)) +
  # add lables
  labs(title = "Deaths per week",
       x = "Week number",
       y = "Deaths",
       col = "Years") +
  # place legend and customise text size to make it all fit
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        text = element_text(size = 7)) + 
  guides(col = guide_legend(nrow = 3)) 

```

## Remarks and additions

As opposed to Excel, R is not "smoothing" the lines between the different data
points for each week. If this is a goal this post on 
[stackoverflow](https://stackoverflow.com/questions/35205795/plotting-smooth-line-through-all-data-points)
might be something for you.

By the time of writing, rendering and finally pushing this post to github, there
were more data available than in the post on reddit. I decided to plot these data
as well.

Since the data have information on age categories, plotting several
graphs is quite easy. Instead of keeping only data for all age groups
(see `filter(Age31December == "10000"`),
we can keep all age groups and use the `facet_wrap()` function from **ggplot2**.
Notice
that I use `scales = "free"` to "zoom" into the different age groups.

```{r}
db_plot2 <- db_cbs %>%
  mutate(year = factor(substr(Periods, start = 1, stop = 4)), # extract year
         wk = factor(as.numeric(substr(Periods, start = 7, stop = 8)))) %>% #extract weeks
        filter(Sex == "T001038", # only keep data for both sexes
               wk %in% 1:52) %>% # only keep weeks 1 through 52
        select(Sex_label, Deaths_1, year, wk, Age31December_label) %>% # keep necessary data only
  mutate(Sex_label = factor(Sex_label))

ggplot(data = db_plot2) +
  # plot lines
  geom_line(aes(x = factor(wk), y = Deaths_1, col = year, group = year)) +
  # add lables
  labs(title = "Deaths per week",
       x = "Week number",
       y = "Deaths",
       col = "Years") +
  # place legend and customise text size to make it all fit
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        text = element_text(size = 7)) + 
  guides(col = guide_legend(nrow = 3)) +
  facet_wrap(~ Age31December_label, scales = "free")
```

Finally, we could also only include the last five years of data to avoid any
discussions of the data not being comparable due to ageing, migration patterns,
advances in vaccination programmes etc.

We can filter data in `ggplot()` directly:

```{r}
ggplot(data = filter(db_plot2, year %in% 2015:2020)) +
  # plot lines
  geom_line(aes(x = factor(wk), y = Deaths_1, col = year, group = year)) +
  # add lables
  labs(title = "Deaths per week",
       x = "Week number",
       y = "Deaths",
       col = "Years") +
  # place legend and customise text size to make it all fit
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        text = element_text(size = 7)) + 
  guides(col = guide_legend(nrow = 1)) +
  facet_wrap(~ Age31December_label, scales = "free")
```

# Conclusion
This post is avoiding any discussion on the usefulness of the data to interpret
the COVID-19 pandemic on purpose. Instead it shows how to easily retrieve
up-to-date data from CBS and plot them with a
few lines of R code. The advantage to this approach (as opposed to plotting in
Excel) is that the graphs are produced
in a transparent and reproducible way and that they can be updated on a weekly
basis (each
time new data is becoming available). This post can serve as a use case for the
**cbsodataR** package or even **ggplot2** (although for the latter there are
much better examples out there).

