---
title: Selecting an appropriate combination of drug dosages with R
author: admin
date: '2019-06-11'
slug: selecting-appropriate-combination-of-drug-dosages-with-r
categories:
  - R
tags:
  - Decision Modelling
image:
  caption: ''
  focal_point: ''
---
Calculating appropriate drug costs is crucial for every health economic evaluation.
I remember the first time I had to calculate drug costs. Back then, working with
Microsoft Excel for a *real* project was excitingly new for me. However,
I got quickly frustrated with this seemingly simple task. The price for a 'simple'
dose of let's say 125mg per patient can be calculated in different ways.
Especially, when I had chose a combination from a plethora of smaller doses. It 
may happen that you may have to combine smaller doses of 2.5mg, 5mg, 15mg, 50mg
or 100mg to derive the needed 125mg. Maybe you would chose for 100mg + 50mg and
discard or reuse the rest. Or you go for 100mg + 15mg + 5mg + 5mg. Many more
options are possible and your decision wouldn't really matter if the price of
larger mg's would be a multiple of the smaller doses. But this is not always the
case. Check out the Dutch prices for **hydroxicarbamide** or **lenalidomide** on
[medicijnkosten.nl](https://www.medicijnkosten.nl) (see Table @\ref(tab:med-price)).

```{r med-price}
data.frame(Name = c(rep("Hydroxycarbamide", 2)))
```

The calculation of drugs costs can be quite daunting since many medications are
often substituted with some kind of 'condition'. Chemotherapies for instance, are
often given depending on the body surface area or per kilogram body weight.

In a deterministic analysis, one would first determine the needed dose

Test

```{r}
# Dosing function

DosingFun <- function(target, combi){
        new.target <- loop.target <- ceiling(target / combi[1]) * combi[1]
        
        i <- 1
        dose <- vector()
        quant <- vector()
        
        
        while(loop.target > 0){
                fills <- floor(loop.target / combi)
                dose[i] <- ifelse(any(fills == 0),
                                  combi[which(fills == 0)[[1]] - 1],
                                  combi[length(combi)])
                
                
                quant[i] <- ifelse(any(fills == 0),
                                   fills[which(fills == 0)[[1]] - 1],
                                   fills[length(fills)])
                
                dosefills <- fills * combi
                
                
                loop.target <- loop.target - last(dosefills[dosefills > 0])
                loop.target
                
                i <-  i + 1
                i
        }
        waste <- new.target - target
        df <- cbind(dose, quant) %>% data.frame()
        
        res <- list(df = df,
                    waste = waste)
        
}

# Test
#x <- DosingFun(target = 133, combi = c(25, 100))
```

